\documentclass[9pt,twocolumn,twoside]{pnas-new}

% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment. 
\templatetype{pnasresearcharticle} 

% Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission
\usepackage{amsthm} 
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{multicol} 
\usepackage{array}
\usepackage{hyperref}

\DeclareMathOperator*{\argmax}{argmax} 
\newtheorem{axiom}{Axiom} 
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture} 


% --------------------------------------------------------------------
% \title{The curiosity trick for solving explore-exploit problems.}
\title{A way around the explore-exploit dilemma}
\author[a,b,1]{Erik J Peterson} 
\author[a,b,c,d]{Timothy D Verstynen} 
\affil[a]{Department of Psychology} 
\affil[b]{Center for the Neural Basis of Cognition} 
\affil[c]{Carnegie Mellon Neuroscience Institute}
\affil[d]{Biomedical Engineering, Carnegie Mellon University, Pittsburgh PA}
\leadauthor{Peterson} 
\authordeclaration{The authors have no conflicts of interest to declare.}

% \equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: Erik.Exists@gmail.com}


% --------------------------------------------------------------------
\begin{abstract}
    Balancing exploration for reward, when exploitation is the long-term goal, is mathematically intractable problem in biology. In this paper, we provide an alternative view that does not depend on exploring to optimize for reward, even though this remains the goal of exploitation. Here the goal of exploration is learning itself. Through theory and experiments we show exploration-as-curiosity leads to a tractable solution to any explore-exploit problem. If, that is, information collection is as important as reward collection over a lifetime. We show that our approach succeeds in naturalistic conditions, when rewards are sparse, deceptive, and non-stationary. We suggest four criteria that might be used to distinguish our approach from other theories.
\end{abstract}
    
\significancestatement{
    It's optimal to be curious.
    % An animal exploring a new environment quickly finds itself with a difficult question: Is it better to keep exploring, or exploit and stick with the known? This problem, the explore-exploit dilemma, is mathematically intractable but common to all animals. When one mathematical problem can’t be solved, it’s often good to find another related problem that can be. In this work we show that when exploration is re-imagined as an open-ended search to learn as much as possible, curiosity, it becomes possible to find simple tractable solutions for all explore-exploit problems, if that is curiosity is just as important as reward for long-term survival. 
}

% \dates{This manuscript was compiled on \today}
% \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
\begin{document} 
\verticaladjustment{-2pt} 
\maketitle

\thispagestyle{firststyle} \ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% --------------------------------------------------------------------
\input{introduction}
\input{results}
\input{discussion}
\input{methods}
\input{acknowledgements}
\bibliography{library,tim_refs} 
\newpage
\input{appendix}
% --------------------------------------------------------------------

\end{document}
